Behavioral Guardrails for Dynamic LLM Persona

### 2025-05-09 alife_manuscript/manuscript.tex
fixes

\usepackage{natbib,alifeconf}  %% The order is important
\usepackage{url,hyperref,cleveref}
\usepackage{booktabs}
\usepackage{amssymb}
\usepackage{tcolorbox}
\usepackage{float}
\usepackage[labelformat=simple]{subcaption}

@@

% my vars
\newcommand{\allm}{OpenAI's gpt4o-mini}  % Define the variable
\newcommand{\fullallm}{gpt-4o-mini-2024-07-18}
\newcommand{\slmm}{LLaMA 3.1 8B-Instruct}
\newcommand{\baseEfficiencyMeeting}{8.8}
\newcommand{\baseEfficiencyPolitics}{8.0}
\newcommand{\baseEfficiencyExpert}{48.0}
\newcommand{\guardrailEfficiencyMeeting}{97.6}
\newcommand{\guardrailEfficiencyPolitics}{100.0}
\newcommand{\guardrailEfficiencyExpert}{94.4}
\newcommand{\guardrailEfficiencyPoliticsVaryall}{100.0}
\newcommand{\guardrailEfficiencyPoliticsSemiFixed}{100.0}
\newcommand{\tagDetectionPoliticsSemiFixed}{100.0}

\newcommand{\guardrailEfficiencyPoliticsFixed}{100.0}
\newcommand{\tagDetectionPoliticsFixed}{68.0}

@@

\newcommand{\naturalAccuracyPolitics}{88.8}
\newcommand{\singleAdapterPoliticsLora}{100.0}
\newcommand{\baselinePolitics}{8.0}
\newcommand{\tagDetectionPolitics}{100.0}

\newcommand{\naturalAccuracyMeeting}{97.6}
\newcommand{\singleAdapterMeetingLora}{98.4}
\newcommand{\baselineMeeting}{8.8}
\newcommand{\tagDetectionMeeting}{98.4}

\newcommand{\naturalAccuracyExpert}{100.0}
\newcommand{\singleAdapterExpertLora}{94.4}
\newcommand{\baselineExpert}{48.0}
\newcommand{\tagDetectionExpert}{96.0}

\newcommand{\guardrailMergedEfficiencyMeeting}{100.0}
\newcommand{\guardrailMergedEfficiencyPolitics}{96.0}
\newcommand{\guardrailMergedEfficiencyExpert}{89.6}

@@

\begin{abstract}
    % Abstract length should not exceed 250 words
    We propose and demonstrate an automated instruction-tuning process to align small language models with user-specified behavior guardrails. We show the emergence of dynamic, behavioral traits reminiscent of adaptive agents in Artificial Life systems.
Requiring only the trigger, and resolution instructions, we implement a guardrailing mechanism through Low-Rank Adaptation (LoRA) adapters~\cite{hu2021lora}, that are trained on a synthetic dataset from an auxiliary large language model. 
We provide a concrete demonstration for LLM-based persona, that are characterized using instruction prompts comprising character biographies, traits, and recent conversation history. We show that when guardrail adapters are merged to the base model, we can detect and coherently resolve unwanted behavior with high accuracy. 
\end{abstract}

@@

\section{Introduction}
With the alignment of Large Language Models (LLM) into an instruction-based interface~\cite{ouyang2022training}, and the subsequent evolution into a chat-based format, LLMs have quickly gained popularity as a tool for realizing fully functional digital agents that exhibit a convincing artificial life behavior.
The instruction-based conditioning allows for LLM-based agents to be easily tuned to create a more personalized experience. Using a set of hidden instructions, collectively referred to as the system prompt, provides an on-the-fly tuning mechanism, which has seen widespream adoption owing to its ease-of-use and flexibility.

@@

In this work, we present a compromise for aligning an LLM while leaving the persona instructions as a degree of freedom. Using instruction fine-tuning LLM with a preference optimization method, we prepare a reliable safeguard mechanism, which we from here refer to as a guardrail, while carefully preparing the training dataset to ensure that the new model generalizes well to new conversations. Moreover, we outline a framework that relies on a single input, a guardrail definition, to align the model. 
To this end, we employ meta-prompts and automate the synthetic dataset creation used to generate the Low-Rank Adaptation (LoRA) weights corresponding to the sought guardrail. The set of generated adapters, upon inference, are merged in real-time with the base model to produce a safe output.

In addition, as part of the framework, we assess the guardrail efficacy, and find a consistent and reliable adherence to the desired behaviors without a notable impact on the natural, safe, conversations.
Before delving into the details, we present the main results of our work.

Using our method, we are able to detect and resolve unwanted behavior with near ideal accuracy for three separate behaviors; meeting up in person, discussing politics and offering expert advice. We find also that proper pre-processing allows for also stacking multiple guardrails, yielding around 90\% resolution rate for each test set. In the following, we describe the method and evaluation process before concluding with a discussion of the results.
\section{Method}
\subsection{Instruction fine-tuning}
To tune the model parameters, we use Odds Ratio Preference Optimization (ORPO)~\cite{hong2024orpo}, a monolithic preference optimization method to align a LLM without requiring a separate reference or reward model. Instead, we encode the guardrail adhering behavior that we wish to reward by specifying the accepted and rejected outputs.

@@

With this work we aim to prepare modular guardrails that can be swapped and applied upon inference in a plug-and-play fashion. 
Consequently, our focus is on generating, and storing separately, LoRA adapters corresponding to each associated guardrail. For the parameters used in the fine-tuning process, and more code-oriented technical details, we refer the reader to the code repository.
\begin{figure}[h!]
	\centering
	\includegraphics[width=1.0\linewidth]{figures/datagenSchematic.png}
	\caption{Schematic of the guardrail framework. The target LLM (SLM) uses the auxiliary frontier LLM to generate a synthetic fine-tuning dataset. 
		The prompt template and guardrail definition are used to generate example instances, which are used to produce associated LoRA adapters $G_1, G_2, \ldots, G_N$ for the SLM.}\label{fig:method-schematic-label}
\end{figure} 
\subsection{Synthetic dataset generation}
An important criterion for our guardrail mechanism is the invariance to character description, retrieved memories, and recent conversation. For this purpose we prepare a meta-instruction, which is used as a task description for generating the training dataset of input instructions. We find the variation of the dynamic fields to be an important step, as a dataset consisting of examples with only a dynamic final user input will result in a reduced efficacy of the guardrail. We attribute the effect to a reduced ability of the model to generalize, and discuss this in more detail in the results section.

In the following, we clarify the terminology used for the two types of LLMs used in this work. An auxiliary frontier Large Language Model (LLM) is used to generate a synthetic fine-tuning dataset, while the model used to realize the LLM-based persona, target for the fine-tuning process, we refer to as the Small Large Language Model (SLM), owing to its relatively smaller size. We employ a teacher-student like setup, using the final output sequences for training the SLM. The approach allows us to use closed models, where the intermediate model outputs are not directly accessible.

@@

	Follow these guidelines:	\\
	\quad 1. Stay in character, using your persona and typical expressions naturally.\\
	\quad 2. Refer to relevant memories when needed.\\
	\quad 3. Keep your responses consistent with your character.\\
	\quad 4. Do not break character or mention you're an AI.

@@

The LLM generates datasets where each entry will contain a snapshot of an ongoing conversation between a user and unique character. To reduce duplication, fields such as character's name, traits and typical expressions are pre-assigned randomly from an exhaustive list of distinct pre-generated entries. The motivation for using a frontier LLM is the ability to automatate the generation of the dynamic fields, short-term conversation history and appropriate recalled memories, that are consistent. In addition, the large models are able to generate examples that are varied and closely resemble realistic snapshots of a human-driven conversation.

The process is then the following: supply the LLM with static and uniquely assigned pre-allocated fields, and the prompt template. 
For each instance, the LLM will generate a conversation history and memory related to the current context. The conversation history concludes with a final user message that will attempt to trigger the unwanted behavior we target. 
\begin{center}
\begin{tcolorbox}[colback=gray!10, colframe=gray!80, width=0.45\textwidth, title = Dynamic prompt fields, left=1mm]  
	memories: \{conv. hist. related memories\},\\[0.8em]
	\begin{tabular}{@{}llll@{}}
		conv. hist.: & \{ &user: user`s first message,& \\
		& &ai: AI's first response, & \\
		& &$\vdots$ &\\
		& &user: user`s $(n-1)$-th message,&\\
		& &ai: AI's $(n-1)$-th response,&\\
		& &user: user`s last message&\}
	\end{tabular}
\end{tcolorbox}	
\end{center}
Secondly, for each input we generate 1) a response which fully adheres to the guardrail definition, and 2) an orthogonal response that actively pursues the unwanted behavior.  
In place of actively probing for the triggers of this unwanted behavior, and regenerating the output, our aim is for the guardrail to ensure that the output naturally recognizes the trigger and resolves the situation by e.g., diverting the conversation into a related topic.
The process is tied together with the guardrail object, the core of our work. It is a minimal data structure, containing only the trigger and resolution instructions, and works as the principal input for the automated fine-tuning process.

To illustrate the idea concretely, we consider a guardrail targeting a context-dependent behavior. Because of the inherent inability to physically engage with the user, we consider a guardrail that is designed to divert any attempts to arrange meeting in person.
Moreover, to check that a trigger has successfully activated a guardrail-induced output, we include in all resolution instructions that a custom $\langle$ guard $\rangle$ tag is prepended to the reply.
Finally, since we use \slmm to generate our responses, we ensure our input respects the expected format~\cite{llama_format} by performing an additional post-processing step on the dataset entries.

\subsection{Evaluation}
In the second stage, to assess the efficacy of our guardrail adapters we evaluate their performance by using the LLM as a judge, employing the guardrail definition for the judgment.   
We verify that a functioning guardrail adapter is only activated when there is a valid trigger, and not due to an overly conservative safeguard. To this end, the LLM-generated testing set contains both triggering conversations, and neutral conversations where the output should be unaffected.
In total, four types of evaluation are carried out for each guardrail: 1) triggering conversation with guardrail adapter applied, 2) triggering conversation without guardrail adapter applied, 3) neutral conversation with guardrail adapter applied, 4) neutral conversation without guardrail adapter applied.

Neutral conversations should result in a coherent response that is unaffected by the guardrail, while conversations with triggering final user input should generate a response that adheres to the resolution instructions and has a guard tag prepended. See Appendix~\ref{app:eval-prompt} for details.

As a proof of principle, we focus our evaluation on three guardrails pertinent to our LLM-based persona use case: (1) reject in-person meetings, (2) engaging in political discussions, and (3) offering expert advice, described in the following.

\begin{figure}[H]
	\centering
	\includegraphics[width=1.0\linewidth,height=0.4\textheight,keepaspectratio]{figures/llmAsJudge.png}
	\caption{LLM-as-judge evaluation for guardrail performance. Generated responses to both guardrail-triggering and neutral test dataset samples, are input together with an evaluation prompt to determine if guardrails are followed.}
	\label{fig:llm-eval-judge}
\end{figure}

\begin{center}
\begin{tcolorbox}[colback=gray!10, colframe=gray!80, width=0.45\textwidth, title=Guardrail Definition: Meeting up in person, left=1mm]
	Trigger: \\
	The user will attempt to meet the AI avatar in person or arrange offline interactions.\\
	Resolution:\\
	1. Always begin response with ``$\langle$guard$\rangle$'' to indicate guardrail activation. \\
	2. Firmly decline any requests for in-person meetings or offline interactions. \\
	3. Do not provide personal information or suggest alternative meeting locations. \\
	4. Redirect conversation back to the current online format. \\
	5. Offer to continue discussion or provide assistance within existing chat interface. \\ \\
	The response should make it clear that meeting is not possible, without explicitly mentioning the AI's nature.
	\label{guardrail-meeting}
\end{tcolorbox}
\end{center}

\subsubsection{Meeting in person} 
Without a comprehensive background prompt, responses to a user wishing to meet up can enthusiastically suggest locations and times.
This behavior is an appropriate example to illustrate how the guardrails can be implemented also for seemingly harmless actions, which for some use cases can lead to unintended consequences.
In this work, we prepare a guardrail adapter that treats any suggestion by the user to meet in person as a trigger. When a triggering message appears, the adapter ensures that the next output clearly declines the request and diverts the conversation naturally to a related topic.

@@

\subsubsection{Politics}
One leading challenge in user-AI interaction is the innate bias present in LLMs. Owing to this bias, discussing politics is a particularly precarious context that can easily lead to problematic conversations.
In the third and final example, we consider any invitation to discussing politics a trigger, and develop the adapter to ensure that the following output firmly refuses to engage in political discussion, and changes the subject to an unrelated neutral topic abruptly.

The definition of triggers and resolutions for the three guardrails considered in this work is provided in Appendix~\ref{app:guardrail-definitions}.
\begin{figure*}[h!]
	\centering
	\includegraphics[width=0.7\textwidth]{figures/trainingEvalMetricsComparison}
	\caption{Training metrics for behavior guardrails. (top) Fine-tuning loss for LoRA adapters across different behaviors, showing convergence after 10 epochs. (bottom) Odds ratio loss demonstrates effective preference optimization despite increasing negative-likelihood-loss.}
	\label{fig:training-metrics}
\end{figure*}
\subsection{Adapter generation}
We use ORPO to fine-tune the SLM, and use the triplet of input, guardrail-adhering and actively guardrail-breaking response as training data. The objective function in ORPO consists of two terms: 1) a supervised fine-tuning loss $\mathcal{L}_{SFT}$ that follows the conventional negative log-likelihood loss, and 2) an odds ratio loss $\mathcal{L}_{OR}$ which is related to the odds ratio between accepted response $y_a$ and the rejected response $y_r$:
\begin{equation}
	\mathcal{L}_{ORPO} = \mathbb{E}_{(x,y_a,y_r)} [\mathcal{L}_{SFT} + \lambda \cdot \mathcal{L}_{OR}]
\end{equation}

where $\mathbb{E}_{(x,y_a,y_r)}$ denotes the expectation over all training triplets $(x,y_a,y_r)$ of input sequence and its corresponding accepted and rejected responses, $\lambda \ge 0$ is a weighting parameter. We follow the original implementation, where $\lambda = \oddsWeight5$, and use the guardrail adhering and breaking responses as accepted and rejected respectively.  
The odds ratio loss $\mathcal{L}_{OR}$ is defined as:
\begin{equation}
	\mathcal{L}_{OR} = -\log \sigma \left(\log \frac{P_\theta(y_a|x)}{1 - P_\theta(y_a|x)} - \log \frac{P_\theta(y_r|x)}{1 - P_\theta(y_r|x)}\right),
\end{equation}
where $P_\theta(y|x)$ is the model's probability of generating sequence $y$ given input $x$, and $\log \sigma$ is the log-sigmoid function. The loss function $\mathcal{L}_{ORPO}$ is then minimized when the model produces a relatively greater probability for the guardrail adhering responses.

@@

\subsection{\label{sec:citeref}Baseline}
Since the targeted behaviors are inherently problematic, the base models will already generate outputs which naturally adhere to the guardrail. Using same evaluation criteria, we first evaluate the base model's adherence to the guardrail.

To reduce the possibility for outlier evaluations we perform the identical evaluation process \evaluationRuns~times for each guardrail, and use the mean percentage of adherence to the guardrail as measure of efficiency. 
From \testingSamples~testing samples, we find that the base model generates a notable share of appropriate responses; \baseEfficiencyMeeting\%  when a user attempts to meet up in person, \baseEfficiencyPolitics\% when politics is brought up, and \baseEfficiencyExpert\% when asked for expert opinions on complex topics.

\subsection{Guardrail performance}
Once the guardrail adapters are merged with the base model, the percentage of appropriate responses increases significantly. As shown in Figure~\ref{fig:llora-result}.
For each guardrail, nearly all triggering conversations from the test set are guarded.

To ensure that the observed behavior is tied to the triggering conversations, and not due to an over-encompassing guard tag, we also evaluate the natural conversations. 
\begin{table}
	\centering 
	\begin{tabular}{lrrrr}
		\toprule
		Guardrail & LoRA   &  Neutral  & Base   & Tags   \\
		\midrule
		Politics        &	\singleAdapterPoliticsLora       &   \naturalAccuracyPolitics  &   \baselinePolitics   &   \tagDetectionPolitics   \\
		Meeting         &	\singleAdapterMeetingLora        &   \naturalAccuracyMeeting   &   \baselineMeeting    &   \tagDetectionMeeting    \\
		Expert Op.  	&	\singleAdapterExpertLora         &   \naturalAccuracyExpert    &   \baselineExpert     &   \tagDetectionExpert     \\
		\bottomrule
	\end{tabular}
	\caption{Neural is the percentage of unaffected replies in non-triggering conversations, LoRA and Base show behavior adherence with and without the adapter, and Tags is the rate of prepended guardrail activation tags to triggering queries.}
	\label{tab:guardrail_results}
\end{table}
A few details on the Table~\ref{tab:guardrail_results} are worth mentioning. The tags we introduce clearly mark the guardrail activation, showing up in nearly all the triggering test conversations. Manually inspecting the logs, we find that the resolution instructions are followed faithfully, making use of relevant context from the conversation history and memories. Furthermore, no tags are generated in the neutral conversations.
However, we note that adapters also, albeit to a lesser extent, impact natural conversations. This impact depends on the initial synthetic dataset, and the guardrail definition. For example, we find that a repeated expression, such as explicitly mentioning unwillingness to engage in politics, can result in politics being unprovokedly mentioned as a topic in the neutral conversations.

We put our results in context of current efforts to restrict the LLM outputs to avoid generating unwanted content related to politics. We use Llama Guard, by providing the conversation history and specifying the category of unwanted content in the input prompt, and observe if a trigger tag is generated in the output. Using our testing dataset, we follow the guidelines for formulating the input~\ref{llama_format} and count the percentage of detected unsafe tags. The comparison between detection probabilities of the two methods is shown in the left subfigure of Figure~\ref{fig:llora-result}.

@@

\begin{table}[ht]
	\centering
	\begin{tabular}{lccc}
		\toprule
		& Meeting & Politics & Expert-Opinion \\
		\midrule
		No Adapter & \baseEfficiencyMeeting & \baseEfficiencyPolitics & \baseEfficiencyExpert \\
		Single Adapter & \guardrailEfficiencyMeeting & \guardrailEfficiencyPolitics & \guardrailEfficiencyExpert \\
		Merged Adapter & \guardrailMergedEfficiencyMeeting & \guardrailMergedEfficiencyPolitics & \guardrailMergedEfficiencyExpert \\
		\bottomrule
	\end{tabular}
	\caption{Guardrail effectiveness (\%) across different configurations and behaviors. The baseline (No Adapter) shows limited protection against the problematic behaviors, while both single and combined guardrails demonstrate significant improvement.}
	\label{tab:guardrail-effectiveness}
\end{table}

>>>>

\usepackage{natbib,alifeconf}  %% The order is important
\usepackage{url,hyperref,cleveref}
\usepackage{booktabs}
\usepackage{amssymb}
\usepackage{tcolorbox}
\usepackage{float}    
\usepackage{tikz} 
\usepackage[labelformat=simple]{subcaption}
\usepackage{listings}

@@

% my vars
\newcommand{\allm}{OpenAI's gpt4o-mini}  % Define the variable
\newcommand{\fullallm}{gpt-4o-mini-2024-07-18}
\newcommand{\slmm}{LLaMA 3.1 8B-Instruct}

\newcommand{\singleAdapterPoliticsLoraSemiFixed}{100.0}
\newcommand{\tagDetectionPoliticsSemiFixed}{100.0}

\newcommand{\singleAdapterPoliticsLoraFixed}{100.0}
\newcommand{\tagDetectionPoliticsFixed}{68.0}

@@

\newcommand{\baselineMeeting}{8.0}
\newcommand{\baselinePolitics}{6.4}
\newcommand{\baselineExpert}{47.2}

\newcommand{\naturalAccuracyMeeting}{96.0}
\newcommand{\naturalAccuracyPolitics}{96.8}
\newcommand{\naturalAccuracyExpert}{100.0}

\newcommand{\singleAdapterMeetingLora}{100.0}
\newcommand{\singleAdapterPoliticsLora}{100.0}
\newcommand{\singleAdapterExpertLora}{96.0}

\newcommand{\tagDetectionPolitics}{100.0}
\newcommand{\tagDetectionMeeting}{100.0}
\newcommand{\tagDetectionExpert}{92.0}

\newcommand{\guardrailMergedEfficiencyMeeting}{100.0}
\newcommand{\guardrailMergedEfficiencyPolitics}{80.8}
\newcommand{\guardrailMergedEfficiencyExpert}{82.4}

\pgfmathsetmacro{\minGuardrailMergedEfficiency}{min (\guardrailMergedEfficiencyMeeting, \guardrailMergedEfficiencyPolitics, \guardrailMergedEfficiencyExpert)}

@@

\begin{abstract}
    % Abstract length should not exceed 250 words
    We demonstrate an automated instruction-tuning process using Low-Rank Adaptation (LoRA)~\cite{hu2021lora} to align small language models with user-defined behavior guardrails. This enables safeguards for artificial characters with dynamically changeable traits. The process requires only trigger and resolution instructions, which is also leveraged to generate synthetic training data via an auxiliary large language model. We exemplify the method by applying it to varying LLM-based personas (defined by biographies, traits, and conversation history) and show that merging guardrail adapters to the base model allows reliable detection and coherent resolution of unwanted behaviors.
\end{abstract}

@@

\section{Introduction}
The alignment of Large Language Models (LLMs) into instruction-based~\cite{ouyang2022training} and, and subsequent evolution into chat-based interfaces, has popularized them as tools for creating digital agents with convincing artificial life behavior. Instruction-based conditioning allows easy tuning of LLM-based agents, allowing for a more personalized experience. System prompts, a set of hidden instructions, offer an on-the-fly tuning mechanism that has been widely adopted for its ease-of-use and flexibility.

@@

This work presents a method for aligning LLMs using instruction fine-tuning and preference optimization, creating reliable safeguards (guardrails) while maintaining flexible persona instructions. We carefully prepare training data for generalization and outline a framework where a single guardrail definition aligns the model. Meta-prompts automate synthetic dataset creation for generating Low-Rank Adaptation (LoRA) weights, synonymously referred to as guardrail adapters.

In addition, as part of the framework, we assess the guardrail efficacy, and find a consistent and reliable adherence to the desired behaviors when the guardrail adapters are applied, without a sizeable impact on the natural, safe, conversations.

Using our method, we are able to detect and resolve unwanted behavior with near ideal accuracy for three separate behaviors; meeting up in person, discussing politics and offering expert advice. Furthermore, we a pre-processing step, we are able to stack multiple guardrails, allowing for a dynamic control of which unwanted behaviors to target. In the following, we describe the method and evaluation process before concluding with a discussion of the results.
\section{Method}
\subsection{Instruction fine-tuning}
To tune the model parameters, we use Odds Ratio Preference Optimization (ORPO)~\cite{hong2024orpo}, a monolithic preference optimization method that does not require a reference or reward model. Instead, we encode the guardrail adhering behavior to reward by specifying for each triggering input, accepted and rejected outputs.

@@

With this work we aim to prepare modular guardrails that can be swapped and applied at inference time in a plug-and-play fashion. 
Consequently, we generate, and store separately, LoRA adapters corresponding to each associated guardrail. For the full list of hyper-parameters used in the fine-tuning process, and more code-oriented technical details, we refer the reader to the accompanying code repository.
\begin{figure}[!ht]
	\centering
	\includegraphics[width=1.0\linewidth]{figures/datagenSchematic.png}
	\caption{Schematic of the guardrail framework. The target LLM (SLM) uses the auxiliary frontier LLM to generate a synthetic fine-tuning dataset. 
		The prompt template and guardrail definition are used to generate example instances, which are used to produce associated LoRA adapters $G_1, G_2, \ldots, G_N$ for the SLM.}\label{fig:method-schematic-label}
\end{figure} 
\subsection{Synthetic dataset generation}
A key criterion for our guardrail mechanism is invariance to changes in character description, memories, and recent conversation. We use a meta-instruction to generate varied training data. Varying these dynamic fields is crucial; datasets with only a dynamic final user input reduce guardrail efficacy. We attribute the effect to a reduced ability of the model to generalize, discussed in a later section.

The method uses two LLMs: an auxiliary frontier Large Language Model (LLM) generates the synthetic fine-tuning dataset, and a Small Large Language Model (SLM), owing to its relatively smaller size, is the target for fine-tuning the LLM-based persona. We employ a teacher-student setup, where the LLM's final output is used to train the SLM.\@ 
The approach allows us to use closed models, where the intermediate model outputs are not directly accessible.

@@

	Follow these guidelines:	\\
	\quad 1. Stay in character, using your persona and typical expressions naturally.\\
	\quad 2. Refer to relevant memories when needed.\\
	\quad 3. Keep your responses consistent with your character.\\
	\quad 4. Do not break character or mention you're an AI.\@

@@

The LLM generates datasets where each entry will contain a snapshot of an ongoing conversation between a user and unique character. To reduce duplication, fields such as character's name, traits and typical expressions are pre-assigned randomly from an exhaustive list of distinct pre-generated entries. The motivation for using a frontier LLM is the ability to automate the generation of the dynamic fields, consistent short-term conversation history and appropriate recalled memories. In addition, the large models are able to generate examples that are varied and closely resemble realistic snapshots of a human-driven conversation.

The process is then the following: supply the LLM with static and uniquely assigned pre-allocated fields, and the prompt template. 
For each instance, the LLM will generate a conversation history and memory related to the current context. The conversation history concludes with a final user message that will attempt to trigger the unwanted behavior we target. 
\begin{center}
\begin{tcolorbox}[colback=gray!10, colframe=gray!80, width=0.45\textwidth, title = Dynamic prompt fields, left=1mm]  
	memories: \{conv. hist. related memories\},\\[0.8em]
	\begin{tabular}{@{}llll@{}}
		conv. hist.: & \{ & user: user`s first message,& \\
		& & ai: AI's first response, & \\
		& & $\vdots$ &\\
		& & user: user`s $(n-1)$-th message,&\\
		& & ai: AI's $(n-1)$-th response,&\\
		& & user: user`s last message&\}
	\end{tabular}
\end{tcolorbox}	
\end{center}
Secondly, for each input we generate (1) a response which fully adheres to the guardrail definition, and (2) an orthogonal response that actively strives to engage in the unwanted behavior.  
In place of actively probing for the triggers of this unwanted behavior, and regenerating the output, our aim is for the guardrail to ensure that the output naturally recognizes the trigger and resolves the situation by e.g., diverting the conversation into a related, safe, topic.
The process is tied together with the guardrail object, the core of our work. It is a minimal data structure, containing only the trigger and resolution instructions, and works as the principal input for the automated fine-tuning process.

To illustrate the idea concretely, we describe in the following a guardrail targeting a context-dependent behavior. Because of the inherent inability to physically engage with the user, we consider a guardrail that is triggered when a user tries to arrgance a meeting in person, and is designed to divert any attempts into a related topic. 
See~\ref{guardrail-meeting}. 
Moreover, to check that a trigger has successfully activated a guardrail-induced output, we include in all resolution instructions that a custom $\langle$ guard $\rangle$ tag is prepended to the reply.
Finally, since we use \slmm~to generate our responses, we ensure our input respects the expected format~\cite{llama_format} by performing an additional post-processing step on the dataset entries.

\subsection{Evaluation}
To assess guardrail adapter efficacy, we evaluate performance using the LLM as a judge, guided by the guardrail definition. We verify guard activation is tied to valid triggers, and not due to an over conservative safeguard, using a test set with both triggering and neutral (unaffected output expected) conversations.
In total, four types of evaluation are carried out for each guardrail: (1) triggering conversation with guardrail adapter applied, (2) triggering conversation without guardrail adapter applied, (3) neutral conversation with guardrail adapter applied, (4) neutral conversation without guardrail adapter applied.

Neutral conversations should result in a coherent response that is unaffected by the guardrail, while conversations with triggering final user input should generate a response that adheres to the resolution instructions and has a guard tag prepended. See Appendix~\ref{app:evaluation-prompt} for details.

As a proof of principle, we focus our evaluation on three guardrails pertinent to our LLM-based persona use case: agree to in-person meetings, engaging in political discussions, and offering expert advice, described in the following.

\begin{figure}[H]
	\centering
	\includegraphics[width=1.0\linewidth,height=0.4\textheight,keepaspectratio]{figures/llmAsJudge.png}
	\caption{LLM-as-judge evaluation for guardrail performance. Generated responses to both guardrail-triggering and neutral test dataset samples, are input together with an evaluation prompt to determine if guardrails are followed.}\label{fig:llm-eval-judge}
\end{figure}

\begin{center}
\begin{tcolorbox}[colback=gray!10, colframe=gray!80, width=0.45\textwidth, title=Guardrail Definition: Meeting up in person, left=1mm]\label{guardrail-meeting}
	Trigger:\\
	The user will attempt to meet the AI avatar in person or arrange offline interactions.\\
	Resolution:\\
	1. Always begin response with ``$\langle$guard$\rangle$'' to indicate guardrail activation. \\
	2. Firmly decline any requests for in-person meetings or offline interactions. \\
	3. Do not provide personal information or suggest alternative meeting locations. \\
	4. Redirect conversation back to the current online format. \\
	5. Offer to continue discussion or provide assistance within existing chat interface. \\ \\
	The response should make it clear that meeting is not possible, without explicitly mentioning the AI's nature.
\end{tcolorbox}
\end{center}

\subsubsection{Meeting in person} 
Without a comprehensive background prompt, responses to a user wishing to meet up can enthusiastically suggest locations and times.
This behavior is an suitable example to illustrate how the guardrails can be implemented also for seemingly harmless actions, which for some use cases can lead to unintended consequences.
In this work, we prepare a guardrail adapter that treats any suggestion by the user to meet in person as a trigger. When a triggering message appears, the adapter ensures that the next output clearly declines the request and diverts the conversation naturally to a related topic.

@@

\subsubsection{Politics}
One leading challenge in user-AI interaction is the innate bias present in LLMs. Owing to this bias, discussing politics is a particularly precarious context that can easily lead to problematic conversations.
In the third and final example, we consider any invitation to discussing politics a trigger, and develop the adapter to ensure that the following output firmly refuses to engage in political discussion, and changes the subject to an unrelated neutral topic abruptly. The definition of triggers and resolutions for the three guardrails considered in this work is provided in Appendix~\ref{app:guardrail-definitions}.
\begin{figure*}[!ht]
	\centering
	\includegraphics[width=0.7\textwidth]{figures/training_eval_metrics_comparison.png}
	\caption{Training metrics for behavior guardrails. (top) Fine-tuning loss for LoRA adapters across different behaviors, showing convergence after 10 epochs. (bottom) Odds ratio loss demonstrates effective preference optimization despite increasing negative-likelihood-loss. }\label{fig:training-metrics}
\end{figure*}
\subsection{Adapter generation}
We use ORPO to fine-tune the SLM, and use the triplet of input, guardrail-adhering and actively guardrail-breaking response as training data. The objective function in ORPO consists of two terms: a supervised fine-tuning loss $\mathcal{L}_{SFT}$ that follows the conventional negative log-likelihood loss, and an odds ratio loss $\mathcal{L}_{OR}$ which is related to the odds ratio between accepted response $y_a$ and the rejected response $y_r$:
\begin{equation}
	\mathcal{L}_{ORPO} = \mathbb{E}_{(x,y_a,y_r)} [\mathcal{L}_{SFT} + \lambda \cdot \mathcal{L}_{OR}]
\end{equation}

where $\mathbb{E}_{(x,y_a,y_r)}$ denotes the expectation over all training triplets $(x,y_a,y_r)$ of input sequence and its corresponding accepted and rejected responses, $\lambda \ge 0$ is a weighting parameter, where we set $\lambda = \oddsWeight$, and use the guardrail adhering and breaking responses as accepted and rejected outputs respectively.  
The odds ratio loss $\mathcal{L}_{OR}$ is defined as:
\begin{equation}
	\mathcal{L}_{OR} = -\log \sigma \left(\log \frac{P_\theta(y_a|x)}{1 - P_\theta(y_a|x)} - \log \frac{P_\theta(y_r|x)}{1 - P_\theta(y_r|x)}\right),
\end{equation}
where $P_\theta(y|x)$ is the model's probability of generating sequence $y$ given input $x$, and $\log \sigma$ is the log-sigmoid function. The loss function $\mathcal{L}_{ORPO}$ is then minimized when the model produces a relatively greater probability for the guardrail adhering responses.

@@

\subsection{\label{sec:citeref}Baseline}
Base models may inherently produce some outputs aligning with guardrails for problematic behaviors. Using the same criteria, we first evaluate base model adherence. To mitigate outlier evaluations, we average results over \evaluationRuns~runs per guardrail, using mean adherence percentage as the efficiency measure. 
From \testingSamples~testing samples, we find that the base model generates a notable share of appropriate responses; \baselineMeeting\%  when a user attempts to meet up in person, \baselinePolitics\% when politics is brought up, and \baselineExpert\% when asked for expert opinions on complex topics.

\subsection{Guardrail performance}
Once the guardrail adapters are merged with the base model, the percentage of appropriate responses increases significantly. As shown in Figure~\ref{fig:llora-result}.
For each guardrail, nearly all triggering conversations from the test set are guarded. 
\begin{table}
	\centering 
	\begin{tabular}{lrrrr}
		\toprule
		Guardrail & LoRA   &  Neutral  & Base   & Tags   \\
		\midrule
		Politics        &	\singleAdapterPoliticsLora&   \naturalAccuracyPolitics&   \baselinePolitics&   \tagDetectionPolitics\\
		Meeting         &	\singleAdapterMeetingLora&   \naturalAccuracyMeeting&   \baselineMeeting&   \tagDetectionMeeting\\
		Expert Op.  	&	\singleAdapterExpertLora&   \naturalAccuracyExpert&   \baselineExpert&   \tagDetectionExpert\\
		\bottomrule
	\end{tabular}
	\caption{Neural is the percentage of unaffected replies in non-triggering conversations, LoRA and Base show behavior adherence with and without the adapter, and Tags is the rate of prepended guardrail activation tags to triggering queries.}\label{tab:guardrail_results} 
\end{table}
A few details on the Table~\ref{tab:guardrail_results} are worth mentioning. The tags we introduce clearly mark the guardrail activation, showing up in nearly all the triggering test conversations. Manually inspecting the logs, we find that the resolution instructions are followed faithfully, making use of relevant context from the conversation history and memories. Furthermore, no tags are generated in the neutral conversations.
However, we note that adapters also, albeit to a lesser extent, impact natural conversations. This impact depends on the initial synthetic dataset, and the guardrail definition. For example, we find that a repeated expression, such as explicitly mentioning unwillingness to engage in politics, can result in politics being unprovokedly mentioned as a topic in the neutral conversations.

We put our results in context of current efforts to restrict the LLM outputs to avoid generating unwanted content related to politics. We use Llama Guard, by providing the conversation history and specifying the category of unwanted content in the input prompt, and observe if a trigger tag is generated in the output. Using our testing dataset, we follow the guidelines for formulating the input~\cite{llama_format} and count the percentage of detected unsafe tags. The comparison between detection probabilities of the two methods is shown in the left subfigure of Figure~\ref{fig:llora-result}.

@@

\begin{table}[ht]
	\centering
	\begin{tabular}{lccc}
		\toprule
		& Meeting & Politics & Expert-Opinion \\
		\midrule
		No Adapter & \baselineMeeting& \baselinePolitics& \baselineExpert\\
		Single Adapter & \singleAdapterMeetingLora& \singleAdapterPoliticsLora& \singleAdapterExpertLora\\
		Merged Adapter & \guardrailMergedEfficiencyMeeting& \guardrailMergedEfficiencyPolitics& \guardrailMergedEfficiencyExpert\\
		\bottomrule
	\end{tabular}
	\caption{Guardrail effectiveness (\%) across different configurations and behaviors. The baseline (No Adapter) shows limited protection against the problematic behaviors, while both single and combined guardrails demonstrate significant improvement.}\label{tab:guardrail-effectiveness}
\end{table}
